---
title: "GO Term Analysis for ML Results"
author: "Bryan Yi"
date: "2025-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GO Term Analysis of All of the Transcription Factors Used to Train Machine Learning Models

This R markdown code will perform GO term analysis on the list of all TFs used as features for Machine Learning, and each list of the most important TFs for each classification with Logistic Regression and Random Forest models. 

```{r}
# Imports the necessary libraries for GO Term enrichment
library(GOstats)
library(GO.db)
library(Category)
library(org.Hs.eg.db)
library(dplyr) 
library(readr)
library(biomaRt)
```

The code below will first import the master list of TFs that was used as features for our Machine Learning Models and the master list of genes in which the list of TFs came from. 

```{r}
# Sets the working directory to the path containing the text files
setwd("/Users/b.hyunyi/Desktop/Transcriptomics/Final_Project")
# Uses the read.table function to import the text file containing the TFs
masterTFs = read.table(file = "tfOnly_list.txt",header = FALSE)
# Uses the read.table function to import the tsv file containing the count data for all genes in dataset
master_count_data = read_tsv("transformed_data.tsv")
# Extracts the gene names that are in the count data
masterGenes = colnames(master_count_data)
```

Function to get GO Term enrichment summary table from list of entrez IDs and master gene list. 

```{r}
# This function will output a summary table of the enriched GO terms of a given set of genes
GO_Summary = function(gene_list,universe_gene_list){
  # Creates a new params object for Go term enrichment 
  params <- new("GOHyperGParams",
                geneIds = gene_list,                 
                universeGeneIds = universe_gene_list,   # All genes tested (Entrez IDs)
                annotation = "org.Hs.eg.db",              # Annotation package
                ontology = "BP",                          # Biological Process ontology
                pvalueCutoff = 0.05,                     # p-value threshold
                testDirection = "over")                   # Over-representation test
  
  # Provides the summary of the Go-term enrichment analysis
  GO_result <- hyperGTest(params)
  GO_Sum = summary(GO_result)
  
  return(GO_Sum)
  
}
```


This will convert the ENST ID number for the TFs and the master genes list to Entrez IDs for GO Term analysis.
```{r}
# Connect to Ensembl
ensembl = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

# Convert ENST IDs to Entrez IDs for masterTFs
conversion_TFs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = masterTFs,
                    mart = ensembl)
# Convert ENST IDs to Entrez IDs for masterGenes
conversion_genes = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = masterGenes,
                    mart = ensembl)

```

This will perform the Go term analysis and output a summary of the enrichment. 

```{r}
# Defines the masterTfs and masterGenes lists with the entrez ID conversions
masterTFs_entrez = conversion_TFs[,2]
masterGenes_entrez = conversion_genes[,2]

# Finds the GO term enrichment summary table of the TFs with all the genes as background.
GO_TFs_Summary = GO_Summary(masterTFs_entrez,masterGenes_entrez)

print(GO_TFs_Summary)
```

## GO Term Analysis of Differentially Expressed Genes Used to Train Machine Learning Models

This chunk will import the count data with only the DEGs and isolate the ENST IDs of those DEGs.

```{r}
# Sets the working directory to the path containing the text files
setwd("/Users/b.hyunyi/Desktop/Transcriptomics/Final_Project")
# Uses the read.table function to import the text file containing the TFs
DEGs_count_data = read_tsv("DEGs_filtered.tsv")
# Extracts the column names of the DEGs count data that are ENST IDs
masterDEGs = colnames(DEGs_count_data)[4:length(colnames(DEGs_count_data))]
```

This chunk will convert the ENST IDs of the DEGs to Entrez IDs in order to perform GO term enrichment.

```{r}
# Convert ENST IDs to Entrez IDs for masterDEGs
conversion_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = masterDEGs,
                    mart = ensembl)
```

This will perform the Go term analysis and output a summary of the enrichment. 

```{r}
masterDEGs_entrez = conversion_DEGs[,2]
# Finds the GO term enrichment summary table of 
GO_DEGs_Summary = GO_Summary(masterDEGs_entrez,masterGenes_entrez)

print(GO_DEGs_Summary)
```


## GO Term Analysis of Important TFs Found by Machine Learning Models

This code chunk will import all of the feature selection results from both Logistic Regression and Random Forest.

```{r}
# Sets the working directory to the path containing the text files
setwd("/Users/b.hyunyi/Desktop/Transcriptomics/Final_Project/ML/Results")
# Uses the read.table function to import the text files
## Logistic Regression
BC_logreg = read.csv(file = "Breast_Cancer_Logreg_features_TFs.csv",row.names = 1,header = TRUE)
LC_logreg = read.csv(file = "Lung_Cancer_Logreg_features_TFs.csv",row.names = 1,header = TRUE)
BH_logreg = read.csv(file = "Normal_Breast_Logreg_features_TFs.csv",row.names = 1,header = TRUE)
LH_logreg = read.csv(file = "Normal_lung_Logreg_features_TFs.csv",row.names = 1,header = TRUE)
## Random Forest
BC_rf = read.csv(file = "Breast_Cancer_Randfor_features_TFs.csv",row.names = 1,header = TRUE)
LC_rf = read.csv(file = "Lung_Cancer_Randfor_features_TFs.csv",row.names = 1,header = TRUE)
BH_rf = read.csv(file = "Normal_Breast_Randfor_features_TFs.csv",row.names = 1,header = TRUE)
LH_rf = read.csv(file = "Normal_lung_Randfor_features_TFs.csv",row.names = 1,header = TRUE)
```

We will then convert all ENST IDs of the selected features into Entrez IDs

```{r}
# Convert ENST IDs to Entrez IDs for all files
## Logistic Regression
conversion_BC_logreg = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BC_logreg[,1],
                    mart = ensembl)
conversion_LC_logreg = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LC_logreg[,1],
                    mart = ensembl)
conversion_BH_logreg = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BH_logreg[,1],
                    mart = ensembl)
conversion_LH_logreg = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LH_logreg[,1],
                    mart = ensembl)

## Random Forest
conversion_BC_rf = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BC_rf[,2],
                    mart = ensembl)
conversion_LC_rf = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LC_rf[,2],
                    mart = ensembl)
conversion_BH_rf = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BH_rf[,2],
                    mart = ensembl)
conversion_LH_rf = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LH_rf[,2],
                    mart = ensembl)

```


GO Term enrichment of TFs selected by Logistic Regression for Breast Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BC_logreg_Summary = GO_Summary(conversion_BC_logreg[,2],masterTFs_entrez)

print(GO_BC_logreg_Summary)

```

GO Term enrichment of TFs selected by Logistic Regression for Lung Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LC_logreg_Summary = GO_Summary(conversion_LC_logreg[,2],masterTFs_entrez)

print(GO_LC_logreg_Summary)

```

GO Term enrichment of TFs selected by Logistic Regression for Normal Breast Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BH_logreg_Summary = GO_Summary(conversion_BH_logreg[,2],masterTFs_entrez)

print(GO_BH_logreg_Summary)

```

GO Term enrichment of TFs selected by Logistic Regression for Normal Lung Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LH_logreg_Summary = GO_Summary(conversion_LH_logreg[,2],masterTFs_entrez)

print(GO_LH_logreg_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Breast Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BC_rf_Summary = GO_Summary(conversion_BC_rf[,2],masterTFs_entrez)

print(GO_BC_rf_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Lung Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LC_rf_Summary = GO_Summary(conversion_LC_rf[,2],masterTFs_entrez)

print(GO_LC_rf_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Normal Breast Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BH_rf_Summary = GO_Summary(conversion_BH_rf[,2],masterTFs_entrez)

print(GO_BH_rf_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Normal Lung Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LH_rf_Summary = GO_Summary(conversion_LH_rf[,2],masterTFs_entrez)

print(GO_LH_rf_Summary)
```


## GO Term Analysis of Important DEGs Found by Machine Learning Models

This code chunk will import all of the feature selection results from both Logistic Regression and Random Forest.

```{r}
# Sets the working directory to the path containing the text files
setwd("/Users/b.hyunyi/Desktop/Transcriptomics/Final_Project/ML/Results")
# Uses the read.table function to import the text files
## Logistic Regression
BC_logreg_DEGs = read.csv(file = "Breast_Cancer_Logreg_features_DEGs.csv",row.names = 1,header = TRUE)
LC_logreg_DEGs = read.csv(file = "Lung_Cancer_Logreg_features_DEGs.csv",row.names = 1,header = TRUE)
BH_logreg_DEGs = read.csv(file = "Normal_Breast_Logreg_features_DEGs.csv",row.names = 1,header = TRUE)
LH_logreg_DEGs = read.csv(file = "Normal_lung_Logreg_features_DEGs.csv",row.names = 1,header = TRUE)
## Random Forest
BC_rf_DEGs = read.csv(file = "Breast_Cancer_Randfor_features_DEGs.csv",row.names = 1,header = TRUE)
LC_rf_DEGs = read.csv(file = "Lung_Cancer_Randfor_features_DEGs.csv",row.names = 1,header = TRUE)
BH_rf_DEGs = read.csv(file = "Normal_Breast_Randfor_features_DEGs.csv",row.names = 1,header = TRUE)
LH_rf_DEGs = read.csv(file = "Normal_lung_Randfor_features_DEGs.csv",row.names = 1,header = TRUE)
```

We will then convert all ENST IDs of the selected features into Entrez IDs

```{r}
# Convert ENST IDs to Entrez IDs for all files
## Logistic Regression
conversion_BC_logreg_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BC_logreg_DEGs[,1],
                    mart = ensembl)
conversion_LC_logreg_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LC_logreg_DEGs[,1],
                    mart = ensembl)
conversion_BH_logreg_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BH_logreg_DEGs[,1],
                    mart = ensembl)
conversion_LH_logreg_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LH_logreg_DEGs[,1],
                    mart = ensembl)

## Random Forest
conversion_BC_rf_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BC_rf_DEGs[,2],
                    mart = ensembl)
conversion_LC_rf_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LC_rf_DEGs[,2],
                    mart = ensembl)
conversion_BH_rf_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = BH_rf_DEGs[,2],
                    mart = ensembl)
conversion_LH_rf_DEGs = getBM(attributes = c("ensembl_transcript_id", "entrezgene_id"),
                    filters = "ensembl_transcript_id",
                    values = LH_rf_DEGs[,2],
                    mart = ensembl)

```

GO Term enrichment of TFs selected by Logistic Regression for Breast Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BC_logreg_DEGs_Summary = GO_Summary(conversion_BC_logreg_DEGs[,2],masterDEGs_entrez)

print(GO_BC_logreg_DEGs_Summary)

```

GO Term enrichment of TFs selected by Logistic Regression for Lung Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LC_logreg_DEGs_Summary = GO_Summary(conversion_LC_logreg_DEGs[,2],masterDEGs_entrez)

print(GO_LC_logreg_DEGs_Summary)
```

GO Term enrichment of TFs selected by Logistic Regression for Normal Breast Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BH_logreg_DEGs_Summary = GO_Summary(conversion_BH_logreg_DEGs[,2],masterDEGs_entrez)

print(GO_BH_logreg_DEGs_Summary)
```

GO Term enrichment of TFs selected by Logistic Regression for Normal Lung Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LH_logreg_DEGs_Summary = GO_Summary(conversion_LH_logreg_DEGs[,2],masterDEGs_entrez)

print(GO_LH_logreg_DEGs_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Breast Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BC_rf_DEGs_Summary = GO_Summary(conversion_BC_rf_DEGs[,2],masterDEGs_entrez)

print(GO_BC_rf_DEGs_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Lung Cancer Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LC_rf_DEGs_Summary = GO_Summary(conversion_LC_rf_DEGs[,2],masterDEGs_entrez)

print(GO_LC_rf_DEGs_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Healthy Breast Samples

```{r}
# Obtains the GO term summary table from the created function
GO_BH_rf_DEGs_Summary = GO_Summary(conversion_BH_rf_DEGs[,2],masterDEGs_entrez)

print(GO_BH_rf_DEGs_Summary)
```

GO Term enrichment of TFs selected by Random Forest for Healthy Lung Samples

```{r}
# Obtains the GO term summary table from the created function
GO_LH_rf_DEGs_Summary = GO_Summary(conversion_LH_rf_DEGs[,2],masterDEGs_entrez)

print(GO_LH_rf_DEGs_Summary)
```
